---
title             : "Equivalence Testing for Psychological Science"
shorttitle        : "Equivalence Testing"

author: 
  - name          : "Daniel Lakens"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Den Dolech 1, IPO 1.33, 5600 MB, Eindhoven, The Netherlands"
    email         : "D.Lakens@tue.nl"
  - name          : "Anne M. Scheel"
    affiliation   : "1"
  - name          : "Peder Isager"
    affiliation   : "1"

affiliation:
  - id            : "1"
    institution   : "Eindhoven University of Technology"

author_note: >
  We would like to thank Courtney Soderbergh for creating the first version of the TOST function to test two independent proportions.

abstract: >
  Psychologists  need to be able to test for the absence of an effect. Using the Two-One-Sided Tests (TOST) procedure, researchers can easily test whether the observed effects are too small to be meaningful. By specifying a smallest effect size of interest (SESOI) researchers test whether observed effects are surprisingly closer to zero, assuming there was an effect the consider meaningful. We explain a range of approaches to determine the SESOI in psychological science, and provide detailed examples of how equivalence tests should be performed and reported. 
  
keywords          : "Equivalence Testing, NHST, power, TOST"
wordcount         : "X"

bibliography      : ["equivalence.bib"]

figsintext        : no
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : yes

lang              : "english"
class             : "man"
output            : papaja::apa6_pdf
---

```{r include = FALSE}
library("papaja")

# papaja:::fetch_zotero_refs(
#   x = "1786197"
#   , bib_name = "equivalence.bib"
#   , API_key = "F7JGNtCxZGT6q87Eeg2zNWiP"
#   , lib_type = "group"
# )
```

Psychologists should be able to falsify predictions. A common prediction in psychological research is that an effect exists in the population that differs from zero. For example, we might predict that priming American Asian women with their Asian identity will increase their performance on a math test compared to women who are primed with their female identity. To be able to falsify this hypothesis, and design a study that allows for strong inferences [@platt_strong_1964], it is important to specify which test result would *disprove* the hypothesis.




An equivalence test can be used to test whether the observed effect is surprisingly small, assuming a meaningful effect exists in the population. The test is a simple variation of the widely used null hypothesis significance tests (NHST). To understand the idea behind equivalence tests, it is useful to realize the null hypothesis we test against can be any numerical value. When we compare two groups, we often test whether the difference between these groups is zero, but we may sometimes want to test against other values than zero. Imagine a researcher who is interested in voluntary participation in a national program to train young infants' motor skills. The researcher wants to test whether more boys than girls are brought into the program by their parents. One could test whether the difference between boys and girls is zero. However, this would ignore the fact that the human sex ratio of newborn boys to girls is not exactly 1:1, and we should not expect 50% of participants to be boys. On average, 103 boys are born for every 100 girls (CIA, 2017), so approximately 50.74% of applicants should be boys, and 49.26% should be girls (with 0.5074/0.4926≈1.03). If boys and girls were exactly equally likely to be brought into the program by their parents, we would not expect a difference of zero, but 1.5% more boys. This can be used as a null hypothesis: Rather than testing against a difference in proportions of 0, the researcher tests against 0.015. 


Even though 0.015 might be a more sensible value for a null hypothesis, our researcher knows that the true sex ratio in their population will differ slightly from that in the entire world. Rather than testing a point null hypothesis (setting $H_0$ to one specific value), the researcher decides to define a range of values around the difference in proportions of 0.015 that can be considered trivially small, even when the true ratio in the population is not exactly 0, or even 0.015.  The researcher can for example test if the difference is smaller than -0.005, or larger than 0.035. This test against two bounds, with $H_0$ as a range rather than one value, is known as a *minimal effects test* (Murphy, Myors, & Wolach, 2014).(Murphy et al. 2014)




However, even if the researcher would consider any difference in percentages smaller than $-0.5 \%$ or larger than $3.5 \%$ a rejection of the null hypothesis, not every difference is large enough to matter in practice. As long as gender differences in the participation rate are not too extreme, there is no need to spend money to address the gender imbalance in the voluntary participation rate, and all money can be spent on the core training program. To test whether the gender difference in participants is *not large enough to matter, the researcher can perform an *equivalence test*. Equivalence tests examine whether the presence of gender differences that are large enough to matter can be rejected. After an extensive discussion with experts, the researcher decides that as long as the difference in proportions is not larger than 6% the researcher can act as if the gender difference is too small to intervene. Given an expected true difference in the population of 0.015, the researcher will tests if the observed difference in the data is smaller than 0.075, and larger than -0.055. If differences more extreme than both these boundary values can be rejected in two one-sided tests, the researcher will conclude statistical equivalence, and no money will be spent on addressing a gender difference in participation.


One can conclude statistical equivalence whenever a 90% confidence interval[a][b][c][d][e][f] around the observed estimate does not contain the smallest effect size of interest (SESOI[g]). Figure 1 illustrates the possible conclusions we can draw from null hypothesis, minimal effect, and equivalence tests. 


Even though equivalence tests are just a small variation of traditional frequentist null-hypothesis significance tests (and test against the SESOI, instead of a value of 0), their use was limited until the availability of user-friendly software to perform the calculations [@lakens_equivalence_2017]. In this article, we provide several examples of equivalence tests which illustrate the procedure. Furthermore, we discuss different approaches to determining the SESOI for psychological research, and provide detailed reproducible examples of how to perform power analyses when designing equivalence tests, and statistical re-analyses of published psychology experiments.




# Justifying the Smallest Effect Size of Interest




Equivalence tests are performed against a value that is considered the smallest effect size of interest (SESOI). The SESOI can sometimes be based on just noticeable differences, which can be objectively determined. Most often, however, it is a subjective decision that varies across individuals and time. The SESOI is ideally based on a cost-benefit analysis. Since both costs and benefits are necessarily subjective, the SESOI will depend on the researcher who designs the study. The goal of setting a SESOI is to clearly justify why designing a study that has a high probability of rejecting effects larger than a specified value contributes to our knowledge base. A SESOI should be chosen such that inferences based on it answer a meaningful question.




##Objective Justifications of a SESOI




An objectively determined SESOI should be based on quantifiable theoretical predictions, such as computational models. Sometimes, the only theoretical prediction is that an effect should be noticeable. In such circumstances, the SESOI can be set based on just noticeable differences. For example, Burriss and colleagues [-@burriss_changes_2015] examined whether women displayed an increase in redness in the face during the fertile phase of their ovulatory cycle. The hypothesis was that a slightly redder skin signals greater attractiveness and physical health, and sending this signal to men yields an evolutionary advantage. This hypothesis requires that the increase in redness can be detected with the naked eye by men. They collected data from 22 women and showed that there was indeed an increase in redness of the facial skin of woman during their fertile period. However, this increase was not large enough to be noticeable with the naked eye by men, thus falsifying their hypothesis. Because the just noticeable difference in redness of the skin can be measured, it is possible to objectively establish the SESOI.




Another example of an objectively determined SESOI can be found in @button_minimal_2015 where the minimal clinically important difference on the Beck Depression Inventory - II was determined by asking 1039 patients when they subjectively felt less depressed (i.e., when they personally noticed an improvement) and relating this to the corresponding difference score on the depression inventory.




##Subjective justifications of a SESOI




We distinguish between three categories of subjective justifications for SESOI. First, researchers can use benchmarks. For example, one might set the SESOI to a standardized effect size of d = 0.5[h][i][j][k], which would allow one to reject effect as large or larger than a 'medium' effect size (Cohen, 1988). Similarly, effect sizes smaller than a Cohen's d of 0.1 are sometimes considered trivially small (Maxwell, Lau, & Howard, 2015). Relying on a benchmark is the weakest possible justification of a SESOI, and should be avoided.




```{r, results="hide"}
library(pwr)
alpha <- 0.05
N <- 100
pwr.t.test(n=N,sig.level=alpha,power=0.33,type="one.sample",alternative="two.sided")$d
```








Second, researchers can determine the SESOI based on previous work in the literature. Ideally, researchers would specify their SESOI in studies they report, but this is not yet common practice. It is thus up to researchers who build on earlier work to decide which effect size is too small to be meaningful, given an earlier study. Simonsohn (2015) recently proposed to set the SESOI to the effect size that an earlier study on the same effect would have 33% power to detect based on the sample size. This is called the 'small telescopes' approach. For example, consider a study where `r N` participants answered a question, which was analyzed with an one-sample t-test. For a two-sided test with an alpha of `r alpha` this test had 33% power to detect a d = `r round(pwr.t.test(n=N,sig.level=alpha,power=0.33,type="one.sample",alternative="two.sided")$d,3)`.




Another justifiable choice for the SESOI we would like to propose here is to use the smallest observed effect size that could have been statistically significant in the original study. In other words, we decide that effects that could not have yielded a *p* < 0.05 in the original study (if this is your preferred alpha threshold) will not be considered meaningful in the replication either, even if effects of this size are statistically significant in the replication study. Based only on the alpha level and the sample size, we can calculate the critical test value (e.g., *t*, *F*, *Z*). This critical test value can also be transformed to a standardized effect size (e.g., $d = t \sqrt { \frac { 1} { n _ { 1} } + \frac { 1} { n _ { 2} } }$), which can thus be interpreted as a *critical effect size*. All observed effect sizes smaller than the critical effect size would not be statistically significant in an original study, given the alpha and sample size of that study. By setting the SESOI to the critical effect size, an equivalence test can reject all observed effect sizes that could have been detected in an earlier study.




```{r, results="hide"}
library(TOSTER)
powerTOSTone(alpha=0.05, statistical_power=0.9, low_eqbound_d=-0.33, high_eqbound_d=0.33)
```




Third[l][m], researchers can set the SESOI based on the resources they have available. The amount of data you can collect limits the inferences you can make. Given a an alpha level and a sample size, researchers can calculate the smallest effect size that one has sufficient power to detect.[n] This approach is suggested here to provide researchers with a way to think about setting the SESOI for a study in a line of research where no quantifiable predictions are made, and where no previous studies have been performed. Observing statistical equivalence for a SESOI based on this approach allows one to reject effects that can be reliably detected with the given sample size. For example, a researcher who plans to perform a two-sided one-sample *t*-test using an alpha of 5%, based on data from 100 observations, has 90% power to detect an effect of d = 0.364. Using equivalence bounds of d = -0.364 and d = 0.364 in an equivalence test allows one to reject effects the study had high statistical power to detect. In this case, concluding equivalence would suggest one can reject effects that can reasonably be detected with N = 100. Whether or not this is interesting depends in part on how typical the chosen sample size is for a research area. One should not expect an equivalence test based on 15 observations testing against a SESOI of d = -1 and d = 1 to substantially increase our knowledge, given that most effects in psychology are substantially smaller than these bounds. But by transparently reporting the effects one can detect and reject, based on the study design, researchers can communicate the information their study contributes. When there are no previous studies to build on, and no theoretical predictions, specifying a SESOI based on the alpha level and a sample size can provide a starting point for a discussion on what a reasonable SESOI is. Nevertheless, the value of setting a SESOI based on this procedure is limited, and the equivalence test does not reject values that are theoretically interesting. 


```{r}
p <- 0.05
N <- 100
crit_d <- abs(qt(p/2, (N*2)-2))/sqrt(N/2)
```


## Example 1: Statistically Equivalent and Not Statistically Different


```{r include = FALSE}
library(TOSTER)


# original study: Banerjee et al., study 1
orig.m1 <- 4.71
orig.m2 <- 5.3
orig.sd1 <- 0.85
orig.sd2 <- 0.97
orig.d <- 0.65
orig.t <- 2.03
orig.p <- 0.049
orig.N <- 40 #group size unknown, therefore equal n assumed
orig.df <- orig.N-2


# replication study: Brandt et al., study 1
rep.m1 <- 4.7857
rep.m2 <- 4.6569
rep.sd1 <- 1.0897
rep.sd2 <- 1.1895
rep.n1 <- 49
rep.n2 <- 51
rep.t <- 0.56
rep.p <- .574
rep.df <- rep.n1 + rep.n2 - 2
rep.d <- 0.11


# Calculate critical effect (d): smallest effect orig. study had the power to detect
t.crit = qt(1-.05/2, (orig.N-2))
d.crit = t.crit*sqrt((1/(orig.N/2)) + 1/(orig.N/2))


Brandt <- TOSTtwo(m1=rep.m1, m2=rep.m2, sd1=rep.sd1, sd2 = rep.sd2, n1 = rep.n1, n2=rep.n2, low_eqbound_d = -d.crit, high_eqbound_d = d.crit, var.equal = FALSE)
```


Banerjee, Chatterjee, & Sinha (2012) reported that participants who had been asked to describe an unethical deed from their past judged the room to be darker than participants who had been asked to describe an ethical deed ($M_{unethical}= `r orig.m1`$, ${SD}_{unethical}= `r orig.sd1`$, $M_{ethical}=`r orig.m2`$, ${SD}_{ethical}=`r orig.sd2`$, $t(`r orig.df`)= `r orig.t`$, $p= `r orig.p`$, $d= `r orig.d`$). A close replication by Brandt, IJzerman, & Blanken (2014) found no significant effect ($t(`r rep.df`)=`r rep.t`$, $p=`r rep.p`$, $d=`r rep.d`$).  The smallest effect size that could have been statistically significant in the original study is $d_{crit}= `r round(t.crit, 2)` \sqrt { \frac { 1} {`r orig.N/2`} + \frac { 1} {`r orig.N/2`} }=`r round(d.crit, 2)`$. Using this as our SESOI for a TOST with Welch's *t*-test for independent samples --- resulting in equivalence bounds of ${\Delta}_L=`r round(-d.crit, 2)`$ and ${\Delta}_U=`r round(d.crit, 2)`$ --- we indeed find that the effect reported by the replication study is statistically equivalent, $t(`r round(Brandt$TOST_df, 2)`)=`r round(min(Brandt$TOST_t1, Brandt$TOST_t2), 2)`$, $p=`r round(max(Brandt$TOST_p1, Brandt$TOST_p2), 3)`$.




```{r, fig.width=6}
TOSTtwo(m1=rep.m1, m2=rep.m2, sd1=rep.sd1, sd2 = rep.sd2, n1 = rep.n1, n2=rep.n2, low_eqbound_d = -d.crit, high_eqbound_d = d.crit, var.equal = FALSE)
```


[^1]: This will typically, although not always, correspond to the effect size the study had $50\%$ power to detect[o]. This procedure will thus result in equivalence bounds that are substantially larger than the ones obtained using the small telescopes approach, which gives the effect size a study had $33\%$ power to detect[p][q][r][s][t][u][v][w].






## Example 2: Statistically Equivalent and Statistically Different


```{r, fig.width=6}
library("TOSTER")
TOSTmeta(ES = 0.06, se = 0.003, low_eqbound_d=-0.1, high_eqbound_d=0.1, alpha=0.005)
TOSTmeta(ES = 0.04, se = 0.002, low_eqbound_d=-0.1, high_eqbound_d=0.1, alpha=0.005)
TOSTmeta(ES = -0.01, se = 0.002, low_eqbound_d=-0.1, high_eqbound_d=0.1, alpha=0.005)
TOSTmeta(ES = -0.01, se = 0.002, low_eqbound_d=-0.1, high_eqbound_d=0.1, alpha=0.005)
TOSTmeta(ES = -0.01, se = 0.002, low_eqbound_d=-0.1, high_eqbound_d=0.1, alpha=0.005)
TOSTmeta(ES = -0.02, se = 0.002, low_eqbound_d=-0.1, high_eqbound_d=0.1, alpha=0.005)
TOSTmeta(ES = -0.02, se = 0.002, low_eqbound_d=-0.1, high_eqbound_d=0.1, alpha=0.005)
TOSTmeta(ES = -0.01, se = 0.003, low_eqbound_d=-0.1, high_eqbound_d=0.1, alpha=0.005)
TOSTmeta(ES = 0.04, se = 0.003, low_eqbound_d=-0.1, high_eqbound_d=0.1, alpha=0.005)
TOSTmeta(ES = 0.06, se = 0.003, low_eqbound_d=-0.1, high_eqbound_d=0.1, alpha=0.005)
```
Hyde, Lindberg, Linn, Ellis, and Williams (2008) report that effect sizes for gender differences in mathematics tests across the 7 million students in the US represent trivial differences, where the authors specify a trivial difference as an effect size smaller than d = 0.1. For example, in grade 3 the difference is d = 0.04, with a standard error of 0.002. When we perform equivalence tests on the meta-analytic effect sizes of IQ difference for grades 2 to 11 (using an alpha level of 0.005 to correct for multiple comparisons), using equivalence bounds of d = -0.1 and d = 0.1, we see that indeed, all effect size estimates are measured with such high precision, we can conclude that they fall within the equivalence bound of d = -0.1 and d = 0.1, and can be considered trivially small, according to the bounds set by the authors. However, note that all of the effects are also statistically different from zero, as one might expect when there is no random assignment to conditions, and samples sizes are huge. This is one way in which the use of equivalence tests improve hypothesis testing procedures in science, by distinguishing between statistical significance and practical significance[x]. 


## Example 3: Not Statistically Equivalent and Not Statistically Different


Moon & Roeder (2014), replicating Shih, Pittinsky, and Ambady (1999), conducted a study to investigate whether Asian-American women would perform better on a maths test when primed with their Asian identity. They observed a slightly reversed difference between the Asian primed group (*n* = 53, *M* = 0.46, *SD* = 0.17) and the control (*n* = 48, *M* = 0.50, *SD* - 0.18) which was not significant, d = -0.23, *t*(96.62) = 1.15, *p* = 0.26, two-sided). A non-significant null-hypothesis test leaves two possible conclusions:


1. The data suggest the absence of a meaningful effect, and we can reject effects large enough to be meaningful.
2. The data are not sensitive enough to tell us whether a meaningful effect is present or absent.


Let us imagine that we consult a teacher about the test, and discover that that grades for this test are set at every 6.25% increase in correct answers (F = 0% to 6.25% ... A+ = 93.75% to 100%). We can decide that we are only interested in test differences so far as they correspond to at least a 1 grade point increase or decrease. Thus, our SESOI becomes a difference in raw scores of 6.25%, or 0.0625.


Using the TOSTtwo function in the TOSTER package in R, we can then calculate the TOST by inserting the summary statistics from Moon & Roeder's (2014) data, assuming an alpha of 0.05, and using +/- the SESOI of 0.0625 as our equivalence bounds.[y][z][aa][ab][ac]


```{r echo=FALSE}


#  We first load the relevant datafile into R
moon.data        <- read.csv("NSNEQ_Moon_Example_data.csv")


#  We then calculate the summary statistics from the data that we need in the TOST calculation below
mean.asian   <- mean(moon.data$accuracy[moon.data$identity_salience == 1])
mean.control <- mean(moon.data$accuracy[moon.data$identity_salience == 2])
sd.asian         <- sd(moon.data$accuracy[moon.data$identity_salience == 1])
sd.control   <- sd(moon.data$accuracy[moon.data$identity_salience == 2])
n.asian          <- sum(moon.data$identity_salience == 1)
n.control        <- sum(moon.data$identity_salience == 2)


#  We can now run the TOST!
#  Performing the TOST using the 'TOSTtwo.raw' function allows us to specify equivalence bounds in raw values
tost <- TOSTER::TOSTtwo.raw(m1  = mean.asian,
                                m2  = mean.control,
                                sd1 = sd.asian,
                                sd2 = sd.control,
                                n1  = n.asian,
                                n2  = n.control,
                                low_eqbound  = -0.0625,
                                high_eqbound =  0.0625,
                                alpha = .05)
```


We find that the TOST is non-significant, t(`r round(tost$TOST_df, 2)`) = `r round(min(abs(c(tost$TOST_t1, tost$TOST_t2))), 2)`, p = `r round(max(tost$TOST_p1, tost$TOST_p2), 2)`. Thus, we cannot reject the possibility that the true effect really is larger than our SESOI.[^footnote] 


[^footnote]:[ad] In this example, the bounds of the TOST was set based on a raw effect of 6.25%. We could have converted this raw effect into a standardized effect (Cohen's d) for the current sample, and we would have gotten the exact same result, only then in Cohen's d (equivalence bounds = +/- `r round(0.0625 / sqrt((sd.asian^2 + sd.control^2) / 2), 2)`). Since this is a replication of a previous study, we could also have chosen calculate the Cohen's d that corresponds to the raw SESOI in the *original* study, and used this as the bounds for the replication. Now we would no longer be answering a question about a raw effect independent of the samples. Instead, we are asking whether we in the replication can reject standardized effects larger than a certain standardized effect in the original study. Since the standard deviations of the original study were only slightly higher than in the replication in this case, Cohen's d, and thus the bounds, would also be only sligthly smaller (bounds = +/- `r round(0.0625 / sqrt((0.17^2 + 0.20^2) / 2), 2)`). As standard deviations of the original study compared to the replication becomes larger, the bounds become more narrow.




## Example 4: Statistically Inferior [ae][af]and Not Statistically Different




Lynott et al. (2014) conducted a study to investigate the effect of being exposed to physically warm or cold stimuli on subsequent judgments related to interpersonal warmth and prosocial behavior (replicating Williams and Bargh, 2008). They observed that 51.2% of participants who received a cold pack (n = 427) opted to receive a reward for themselves, while 57.9% (n = 434) of participants who received a warm pack did the same. Calculating a z-test for the difference between the proportions, this effect is statistically significant (Diff = -0.067, z = -1.98, p = 0.024[ag]). Having obtained a significant result, we are left with two alternatives:


1. The difference is small enough that even though it is *statistically* significant, we can consider it *practically* equivalent to zero.
2. The difference is substantial enough that we cannot rule out that it is also practically significant.


In this case, it is not really clear what an objective cirterion for the smallest effect size of interest could be. However, since this is a replication, we could decide that we do not care about an effect if the difference is smaller than the smallest effect that the original study could detect. This essentially means that we use the critical z value (~1.96 in a two-tailed test with an alpha of 0.05) as our bounds. To figure out what difference corresponds to a critical z in the original study, we multiply critical z with the standard error:


```{r}
se         <- sqrt((0.75 * (1 - 0.75)) / 26 + (0.46 * (1 - 0.46)) / 27)  # Calculate the (unpooled) standard error
crit.z <- qnorm(1 - (0.05 / 2))  # Determine the critical z for a two-tailed test with alpha = 0.05


crit.diff <- crit.z * se  # Convert z to the critical difference


print(crit.diff)
```


Having determined the SESOI, we may also decide that since this is a replication, we are only interested in whether the effect of receiving a cold pack is significantly *lower* than the smallest detectable effect of the original study. In this case, we can still use a TOST to analyse the data, but we only need to consider the test against the upper bound because an effect with CIs crossing the lower bound can still be *lower* than the original effect. 


```{r}


download.file(url = "https://raw.githubusercontent.com/Lakens/EquivalenceTestingForPsychologicalScience/master/TOSTtwo.prop.R", destfile = "TOSTtwo.prop.R")
source(file = "TOSTtwo.prop.R")


#  We first load the data and exclude cases to only include cases that did not meet the authors' exclusion criteria.
lynott.data <- read.csv("NSEQ_Lynott_example_data.csv")
lynott.data <- lynott.data[lynott.data$include1 == 1,]  


#  We then calculate the summary statistics from the data that we need in the TOST calculation below
n.hot            <- sum(lynott.data$packtype == 1)
n.cold           <- sum(lynott.data$packtype == 0)
selfish.hot  <- sum(lynott.data$selfish[lynott.data$packtype == 1] == 1, na.rm = TRUE) / n.hot
selfish.cold <- sum(lynott.data$selfish[lynott.data$packtype == 0] == 1, na.rm = TRUE) / n.cold
 
#  We can now run the TOST!
tost <- TOSTtwo.prop(prop1 = selfish.cold,
                         prop2 = selfish.hot,
                         n1        = n.cold,
                         n2        = n.hot,
                         low_eqbound  = -crit.diff,
                         high_eqbound = crit.diff,
                         alpha = 0.05)


```


We find that the test against the upper bound is `r ifelse(tost$TOST_p2 < tost$alpha, "significant", "non-significant")`, $z =$ `r round(tost$TOST_z2, 2)`, $p =$ `r ifelse(round(tost$TOST_p2, 3) >= 0.001, paste("=", round(tost$TOST_p2, 2)), "< 0.001")`. Thus, we can conclude that the statistically non-significant effect is also statistically inferior to our SESOI. 
## Example 5: Not Statistically Equivalent and Statistically Different
Example 5: Correlations[ah]? 


Would be nice to have 1 bounds based on effect sizes you have power to detect. And one on just noticeable differences. This: http://wiki.ihe.net/images/e/e5/Measurement_of_pain.pdf suggests that  this: Berry, H., Huskisson, E. C. Clin. Trials J. 1972, 9, 13.
12. Hardy, J. D., Wolff, H. G., Goodell, H. Pain Sensations and
Reactions. Baltimore, 1952. shows that a VAS for pain can be divided into 20 just noticeable differences' Or we use  






#Adressing Trello cards




For the example where the results are inconclusive, explain how adding more data, will, assuming the true effect is 0, lead to a smaller confidence interval, and thus will lead to an informative test. However, with increasing data, the effect could also slightly increase, to above or at the SESOI level, and then the effect will be significant and not equivalent. With massive amounts of data, many results will be significant and equivalent.
























# Discussion


The given examples and figures XXX to YYY illustrate that equivalence testing is a relatively simple and flexible technique: As long as we can calculate a confidence interval around a mean effect, we can compare it to one or several bounds. In that sense, the result of an equivalence test can be obtained by mere visual inspection of the CI (Seaman[ai] & Serlin, 1998; Tryon, 2001). This ease of use might tempt researchers to set equivalence bounds after looking at their data.
For example, we showed above that Moon & Roeder's replication of the identity-priming study by Shih et al. (1999; example 3) does not exclude effects that are large enough to affect a student's final grade on a test. However, if we look at the $90 \%$ CI (figure XXX), we can see that the data would have been statistically equivalent if the bounds had been based on a slightly larger SESOI, e.g. a mean difference of 0.1.
It is important to point out that this strategy --- setting equivalence bounds *based on the obtained results* --- may be considered as "boundary hacking": Regardless of the underlying true effect, any finite CI will exclude some values, which means that this procedure will lead to capitalization on chance.


In order to avoid fooling ourselves more often than $\alpha \%$ of the time, equivalence bounds should be specified before results are known (ideally as part of a preregistration; cf. CONSORT guidelines, Piaggio et al., 2006). A perhaps helpful guiding question may be: "Will testing against these bounds yield interesting information in the long run?" In most cases[aj],




#Raw[ak] versus standardized equivalence bounds
 
Equivalence bounds can be set in terms of standardized effect sizes, or raw differences. For example, the SESOI might be a raw mean difference of 0.5 points on a 7-point scale, or a Cohen’s d of 0.5. The main difference is that equivalence bounds set in raw differences are independent of the standard deviation, while equivalence bounds set as standardized effects are dependent on the standard deviation (since they are calculated as [FORMULA (X1-X2)/SD]. The observed standard deviation randomly varies across samples. In practice, this means that when you use standardized differences as bounds (e.g., d = 0.35) the equivalence test depends on the standard deviation in the sample. The equivalence bound for a raw mean difference of 0.5 equals a standardized equivalence bound of d = 0.5 when the standard deviation is 1, but a standardized equivalence bound of d = 1 then the standard deviation is 0.5.


Both raw equivalence bounds and standardized equivalence bounds have specific benefits and limitations (for a discussion, see Baguley, 2009). When raw mean differences are meaningful and of theoretical interest, it makes sense to set equivalence bounds based on raw effect sizes. When the raw mean difference is of less theoretical importance, or different measures are used across research lines, it is often easier to set equivalence bounds based on standardized differences. Researchers should realize that equivalence bounds based on raw differences or standardized differences ask slightly different questions, and justify their choice for an equivalence bound. When setting equivalence bounds based on earlier research, such as in replication studies, equivalence bounds based on raw or standardized differences would ideally give the same result, and large differences in standard deviations between studies are as important to interpret as large differences in means.










\newpage




# References




Button, K. S., Kounali, D., Thomas, L., Wiles, N. J., Peters, T. J., Welton, N. J., . Lewis, G. (2015). Minimal clinically important difference on the Beck Depression Inventory - II according to the patient's perspective. Psychological Medicine, 45(15), 3269-3279. https://doi.org/10.1017/S0033291715001270




Burriss, R. P., Troscianko, J., Lovell, P. G., Fulford, A. J. C., Stevens, M., Quigley, R., . Rowland, H. M. (2015). Changes in Women's Facial Skin Color over the Ovulatory Cycle are Not Detectable by the Human Visual System. PLOS ONE, 10(7), e0130093. https://doi.org/10.1371/journal.pone.0130093


Lakens, D. (2017). Equivalence Tests: A Practical Primer for t Tests, Correlations, and Meta-Analyses. Social Psychological and Personality Science, 8(4), 355–362.


\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}
