---
title: "NSNEQ_Moon_Example"
author: "Peder M Isager"
date: "2017 M10 16"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Example of an effect that is neither statistically significant nor equivalent

Moon & Roeder (2014), replicating Shih, Pittinsky, and Ambady (1999), conducted a study to investigate whether Asian-American women would perform better on a maths test when primed with their Asian identity. They observed a slightly reversed difference between the Asian primed group (N = 53, M = 0.46, SD = 0.17) and the control (N = 48, M = 0.50, SD - 0.18) which was not significant, d = -0.23, t(96.62) = 1.15, p = 0.26, two-sided). Having failed to obtain a significant result, we are left with two alternatives:

1. The data indicate the absence of a meaningful effect, and we can accept the null hypothesis.
2. The data are not sensitive enough to tell us whether a meaningful effect is present or absent. 

In order to distinguish between these two hypotheses, we first need to define a SESOI (and thus, our equivalence bounds). Let us imagine that we consult a teacher about the test, and discover that that grades for this test are set at every 6.25% increase in correct answers (F = 0% to 6.25% ... A+ = 93.75% to 100%). We can decide that we are only interested in test differences so far as they correspond to at least a 1 gradepoint increase or decrease. Thus, our SESOI becomes 6.25%, or 0.0625. 

Using the TOSTtwo function in the TOSTER package in R, we can then calculate the TOST by insterting the summary statistics from Moon & Roeder's (2014) data, assuming an aplha of 0.05, and using +/- the SESOI of 0.0625 as our equivalence bounds. 

```{r}

#  We first load the relevant datafile into R
moon.data    <- read.csv("NSNEQ_Moon_Example_data.csv") 

#  We then calculate the summary statistics from the data that we need in the TOST calculation below
mean.asian   <- mean(moon.data$accuracy[moon.data$identity_salience == 1])
mean.control <- mean(moon.data$accuracy[moon.data$identity_salience == 2])
sd.asian     <- sd(moon.data$accuracy[moon.data$identity_salience == 1])
sd.control   <- sd(moon.data$accuracy[moon.data$identity_salience == 2])
n.asian      <- sum(moon.data$identity_salience == 1)
n.control    <- sum(moon.data$identity_salience == 2)

#  We can now run the TOST!
#  Performing the TOST using the 'TOSTtwo.raw' function allows us to specify equivalence bounds in raw values
tost <- TOSTER::TOSTtwo.raw(m1  = mean.asian,
                            m2  = mean.control,
                            sd1 = sd.asian,
                            sd2 = sd.control,
                            n1  = n.asian,
                            n2  = n.control,
                            low_eqbound  = -0.0625,
                            high_eqbound =  0.0625,
                            alpha = .05)
```

We find that the TOST is non-significant, t(`r round(tost$TOST_df, 2)`) = `r round(min(abs(c(tost$TOST_t1, tost$TOST_t2))), 2)`, p = `r round(max(tost$TOST_p1, tost$TOST_p2), 2)`. Thus, we cannot reject the possibility that the true effect really is larger than our SESOI.


